---
title: "How to create an R package and use Git & Github for version control"
subtitle: "***"
author:  
  - Alex Chen
  - <small>EDIT</small>
date: "<small>`r format(Sys.time(), '%d %B %Y')`</small>"
output:
  revealjs::revealjs_presentation:
    incremental: true
    highlight: pygments
    theme: serif
    center: true
    transition: fade
    css: custom.css
    self_contained: false
    reveal_plugins: ["zoom","notes"]

---
## Link to the slides

[http://tinyurl.com/zzlzs2u](http://tinyurl.com/zzlzs2u)

## Advantages of building an R package

- Use functions and turn hundreds lines of code to a single or few lines of code. 

- Orgainze your R script, data, and documentation in a well-defined structure.  

- Collaborate and share your methods with others. Editing history traceable via git control.  

- Use 'R CMD check' to validate your R script.  

- Publishable work  

<div class="notes">
You may wonder why I would create my own package and what I can benefit from. Actually, there're quite a few advantages.
First, a package basically comprises a set of functions, and you can pack your hundred lines    
Second, when you're building your own package, it kind of forces you to organize...
Third, creating a package actually makes it easier to collaborate and share...
Fourth, it's probably not the attractive one, but you can use R built-in function to validate your R script
Lastly, an R package itself can be a publishable work.
</div>

## R package structure 

**A minimal package should include:**

- R functions (saved in .R files)

- Metadata (DESCRIPTION file)

- Namespace (NAMESPACE file generated via roxygen2; connect your package to others)

- Documentation (generated via roxygen2; help page)

- Vignette (user guide)

<div class="notes">
Here are some basic components of an R package. First is the R functions...
The most important content in an R package in the functions, so we will create an R package by starting to write R functions.
</div>
 
## Writing R functions can save your time down the road

```{r, echo=FALSE, fig.height=5, warning=F}
library(ggplot2);library(reshape2)
Y.axis <- c(0:11)
fun.user <- c(0:11)
fun.writer <- c(0,3,4,5,5.5,6,6,6,6,6,6,6)
df <- as.data.frame(cbind(Y.axis, fun.user, fun.writer))
dfn <- melt(df, id="Y.axis")
ggplot(dfn, aes(x=Y.axis, y=value, colour=variable))+
  geom_smooth(se=F)+
  ylab("Time")+xlab("Data size / complexity")+
  theme_bw()+
  theme(legend.position="none", axis.ticks=element_blank(), axis.text=element_blank(),
        axis.title=element_text(size=22))+
  annotate("text", x=7.3, y=10, label="non-function writer", size=9)+
  annotate("text", x=9, y=5.3, label="function writer", size=9)

```

<div class="notes">
Here is the plot to show you the difference between a function writer and non-function writer. If you are not a function writer, the time your spend on analyzing data or crafting your code tends to increase as your data size and complexity increase. However, if you'r a function writer, you may spend a little more time at the beginning, but it will save your time down the road. 
</div>

## How to write a function

```{r, eval=F}
Your_Function_Name <- function( Your_Variables ){
  #Do something here
}

```
. . . 

Fecundity = a x Length ^b^

. . . 

```{r, message=FALSE, warning=FALSE}
fecundity <- function(L){
  y <- 0.002*(L^3)
  return(y)
}
```
. . . 

```{r}
fecundity(36)
```

## How to write a function

Fecundity = a x Length ^b^

```{r, message=F}
fecundity <- function(L, a, b){
  y <- a*(L^b)
  return(y)
}
```
. . .

```{r}
fecundity(L=36, a=0.002, b=3)
```
. . . 
```{r}
fecundity(L=36, a=0.2, b=3)
```

## Make your function useful

Use conditional statement to choose an analytical method

```{r }
myFunction <- function(method="A"){
  
  if(method=="A"){
    #Analyze data in this way
  
  }else if(method=="B"){
    #Analyze data in that way
  
  }else if(method=="C"){
    #Analzye data in that way
    
  }
}
```

## Start an R package with git version control

Your R package is built in your computer; 
a copy can be saved in the cloud. They're syncronized via Git. 

|    | Local | Cloud |
|:--:|:-----:|:-----:| 
| storage space | Your PC | Github.com  | 
| software | RStudio | Git | 

## Start an R package

- Install R, RStudio, and [Git](https://git-scm.com/downloads) in your computer

- Create a Github account [www.github.com](http://github.com)

- [Create a new repository](pics/ScreenShot-Github-create-a-respository.png) / [Copy the repository's link](pics/ScreenShot-Github-copy-link.png)

- Create an R project from RStudio
    1. Open a new project
    2. [Select version control](pics/ScreenShot-start-package-1-select-version-control.png)
    3. [Select Git](pics/ScreenShot-start-package-2-select-Git.png)
    4. [Paste URL](pics/ScreenShot-start-package-3-paste-URL.png)

- Install library `devtools` and `roxygen2`

- Set up R configuration

## R configuration setup

1. Select **Build** and then **Configure Build Tools** on [menu bar](pics/ScreenShot-Rconfigure-1-ConfigureBuildTools.png)

2. Select **Package** in the pop-up window

3. Check and configure [Roxygen options](pics/ScreenShot-Rconfigure-2-RoxygenOptions.png)

4. Create folders **R** and **man**

5. Create a text file named **DESCRIPTION**

## Edit **DESCRIPTION** file

Include the following information:

```{r, eval=FALSE}
Package: MyRpackage
Version: 1.0
Title: This is title of MyRpackage
Author: Alex Chen
Maintainer: Alex Chen <chen.1735@osu.edu>
Description: A brief description about MyRpackage
Depends: R (>= 3.1.0)
License: GPL (>= 2)
Imports: [Put other packages here if needed]

```

More details about DESCRIPTION

[R official site](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-DESCRIPTION-file)

[by Hadley Wickham](http://r-pkgs.had.co.nz/description.html)


## Write R function files

Create function files (e.g., fecundity.R) \
and save them in the **R** folder

```{r, eval=F}
#' A title of your function
#'
#' A short description about the function.
#' @param L fish length
#' @param a constant a
#' @param b constant b
#' @return This function returns fish fecundity
#' @export
#' @examples
#' fecundity(36, 0.2, 3)

fecundity <- function(L, a, b){
  y <- a*(L^b)
  return(y)
}

```

We use [**roxygen2**](https://cran.r-project.org/web/packages/roxygen2/README.html) to generate functions' documentation.

## Build your package in R

- Click **Build & Reload** under [**Build** menu](pics/ScreenShot-BuildReloadPackage-1-click-Build-Reload.png)

- Fix any issue if build not successful

- Make sure to rebuild if any change was made

- Try `?fecundiy` to see if help page shows up 

## Sync with Github repository

- Make sure the [Git program](https://git-scm.com/downloads) was installed.

- Click **Commit** under **Git** [menu](pics/ScreenShot-GitSync-1-click-commit-under-git.png)

- Check files and add commit message and then hit [**commit**](pics/ScreenShot-GitSync-2-check-commit-message.png)

- Close commit message [window](pics/ScreenShot-GitSync-3-git-commit.png) 

- Click [**Push**](pics/ScreenShot-GitSync-4-git-push.png) button to upload files to Github repository

- Go to your Github site and see if files are updated

## Exercise 

- Go ahead and create a second R function

- Rebuild your package locally, and then upload the file to your Github

- Install the package of a person next to you. Use the following code:

. . . 

```{r, eval=F}
library(devtools)

install_github("ACCOUNT/XPACKAGE")

```

## Handle identical function names

- Import the library: `library(XPACKAGE)`

- Type `?fecundity` in R console (There should be two *fecundity* functions associated with different packages)

- Type `?XPACKAGE::fecundity` to see the help page

- Click **Index** at the bottom to see the second function

- Run the second function to see if it works 

## Further resources

[One-page tutorial](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/) (by Hilary Parker)

[R package primer](http://kbroman.org/pkg_primer/) (by Karl Broman)

[R packages](http://r-pkgs.had.co.nz/) (by Hadley Wickham)

[R official document](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)

### ~~~

### Happy coding!
